---
title: "Capone et al. — When cooler is better: steep thermal sensitivity coexists with sustained respiration under acute warming in Galápagos octocorals"
subtitle: "Complete analysis pipeline: data processing, analyses, and figures"
author: "Haley E. Capone"
date: "2026"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5.5)
```

# Overview

This script accompanies the manuscript:

> **"When cooler is better: steep thermal sensitivity coexists with sustained respiration under acute warming in Galápagos octocorals"**
> Capone HE, Silva IN, Brandt ME, Bruno JF
> Submitted to *Marine Ecology Progress Series* (MEPS-2025-12-043)

The pipeline proceeds in five parts:

1. **Data processing** — Raw O₂ timeseries → respiration rates (via LoLinR)
2. **Primary analyses** — Q₁₀, Arrhenius activation energy (E), rate-by-temperature ANOVAs, ANOVA assumption diagnostics, site temperature comparison
3. **Supplementary analyses** — TPC model selection (peaked vs. monotonic, 5 candidate models)
4. **Figures** — Main text (Figs. 1–3) and supplementary (Fig. S1)
5. **Table export** — Tables S1–S3 and summary statistics

```{r packages}

# packages
library(tidyverse)
library(ggplot2)
library(cowplot)
library(minpack.lm)
library(ggtext)
library(car)

# For raw data processing (Part 1 only):
# library(LoLinR)
# library(Rmisc)
# library(stringr)
```

```{r constants}
# Boltzmann constant (eV/K)
k_B <- 8.617e-5

# Species order and aesthetics (used throughout)
sp_order  <- c("Pdarw", "Mplan", "Bgran", "Tcocc")
sp_colors <- c("Pdarw" = "#E8706A", "Mplan" = "#9B7BCA",
               "Bgran" = "#D4C98E", "Tcocc" = "#E88A28")
sp_names  <- c("Pdarw" = "P. darwinii", "Mplan" = "M. plantaginea",
               "Bgran" = "B. grande", "Tcocc" = "T. coccinea")
```


---

# Part 1: Data processing (raw O₂ → rates)

This section processes raw PreSens O₂ timeseries files into per-replicate
respiration rates using LoLinR (Olito et al. 2017). The output is
`Photo.R_full.csv`, which is the input for all subsequent analyses.

**Note:** This section is provided for reproducibility but is computationally
intensive (~289 files). The processed output (`Output/Photo.R_full.csv`) is
included in the repository. To skip to analyses, proceed to Part 2.

```{r set-working-directory}
dir.create("Output", showWarnings = FALSE)
dir.create("Figures", showWarnings = FALSE)
```

## 1.1 Load metadata

```{r load-metadata}
Sample.Info <- read.csv(file = "Cnidarians_metadata.csv", header = TRUE)

# Convert time columns
Sample.Info$Start.time <- as.POSIXct(Sample.Info$Start.time,
                                      format = "%H:%M:%S", tz = "")
Sample.Info$Stop.Time  <- as.POSIXct(Sample.Info$Stop.Time,
                                      format = "%H:%M:%S", tz = "")
```

## 1.2 LoLinR slope extraction

This loop reads each raw O₂ CSV, thins the data, and extracts the
best-fit linear slope using the LoLinR package.

```{r lolinr-extraction, eval=FALSE}
# List all raw O2 files
path.p <- normalizePath(getwd())
file.names      <- basename(list.files(path = path.p, pattern = "csv$",
                                        recursive = TRUE))
file.names.full <- list.files(path = path.p, pattern = "csv$",
                               recursive = TRUE)

# Initialize output dataframe
Photo.R <- data.frame(matrix(NA, nrow = length(file.names), ncol = 4))
colnames(Photo.R) <- c("ID", "Intercept", "umol.L.sec", "Temp.C")

# Extract slopes
for (i in 1:length(file.names.full)) {
  
  FRow <- which(Sample.Info$ID == strsplit(file.names[i], '.csv'))
  
  Photo.Data1 <- read.csv(file.path(path.p, file.names.full[i]),
                            skip = 1, header = TRUE)
  Photo.Data1 <- Photo.Data1[, c("Time", "Value", "Temp")]
  Photo.Data1$Time <- as.POSIXct(Photo.Data1$Time,
                                   format = "%H:%M:%S", tz = "")
  Photo.Data1 <- na.omit(Photo.Data1)
  
  # Trim start/end noise
  n <- nrow(Photo.Data1)
  Photo.Data1 <- Photo.Data1[15:(n - 3), ]
  n <- nrow(Photo.Data1)
  Photo.Data1$sec <- 1:n
  
  rename <- sub(".csv", "", file.names[i])
  
  # Thin data for LoLinR
  Photo.Data.orig <- Photo.Data1
  Photo.Data1 <- thinData(Photo.Data1, by = 15)$newData1
  Photo.Data1$sec <- as.numeric(rownames(Photo.Data1))
  Photo.Data1$Temp <- NA
  Photo.Data1$Temp <- thinData(Photo.Data.orig,
                                 xy = c(1, 3), by = 15)$newData1[, 2]
  
  # LoLinR regression
  Regs <- rankLocReg(xall = Photo.Data1$sec, yall = Photo.Data1$Value,
                      alpha = 0.5, method = "pc", verbose = TRUE)
  
  Photo.R[i, 2:3] <- Regs$allRegs[1, c(4, 5)]
  Photo.R[i, 1]   <- rename
  Photo.R[i, 4]   <- mean(Photo.Data1$Temp, na.rm = TRUE)
}

write.csv(Photo.R, 'Output/Photo.R.csv')
```

## 1.3 Normalize rates

```{r normalize-rates, eval=FALSE}
Photo.R <- read.csv('Output/Photo.R.csv')
Photo.R <- na.omit(Photo.R)

# Join with metadata
Photo.R_full <- left_join(Photo.R, Sample.Info, by = "ID")

# Convert volume to liters
Photo.R_full$Volume <- Photo.R_full$Volume / 1000

# Calculate μmol s⁻¹ (accounts for chamber volume)
Photo.R_full$umol.sec <- Photo.R_full$umol.L.sec * Photo.R_full$Volume

# Factor conversion
Photo.R_full <- Photo.R_full %>%
  mutate_if(sapply(., is.character), as.factor)
Photo.R_full$Blank <- as.factor(Photo.R_full$Blank)

# Calculate and subtract blank rates
photo.blnk <- aggregate(umol.sec ~ ID * Temp.Cat * Light_Dark * Run * Blank,
                          data = Photo.R_full, mean)
photo.blnk <- photo.blnk[photo.blnk$Blank == 1, ]
photo.blnk$Blank <- NULL
colnames(photo.blnk)[4] <- 'blank.rate'

Photo.R_full.blnk <- left_join(Photo.R_full, photo.blnk)
Photo.R_full.blnk$umol.sec.corr <- Photo.R_full.blnk$umol.sec -
                                      Photo.R_full.blnk$blank.rate

# Normalize to AFDW (μmol g⁻¹ h⁻¹)
Photo.R_full.blnk$umol.cm2.hr <- (Photo.R_full.blnk$umol.sec.corr * 3600) /
                                    Photo.R_full.blnk$AFDW
Photo.R_full.blnk <- Photo.R_full.blnk[Photo.R_full.blnk$Blank == 0, ]

# Assign rate type and individual ID
Photo.R_full.blnk$rate.type <- ifelse(Photo.R_full.blnk$Light_Dark == 'Light',
                                        "NP", "R")
Photo.R_full.blnk$individual.ID <- str_split(Photo.R_full.blnk$ID,
                                               ".{1}$", simplify = TRUE)[, 1]

# Make respiration positive
Photo.R_full.blnk$umol.cm2.hr <- -Photo.R_full.blnk$umol.cm2.hr

# Extract dark respiration only
Rdark <- subset(Photo.R_full.blnk, rate.type == "R") %>%
  dplyr::select(Temp.Cat, ID, Species, umol.cm2.hr, Chamber.Channel,
                Run, individual.ID)

# Add mean temperature
sumE <- summarySE(Photo.R_full.blnk,
                   measurevar = "Temp.C",
                   groupvars = c("Run", "individual.ID"))
Rdark$Temp.C <- sumE$Temp.C

# Log-transform (ln)
Rdark$log.rate <- log(Rdark$umol.cm2.hr)

# Save processed data
write.csv(Rdark, 'Output/Photo.R_full.csv', row.names = FALSE)
```


---

# Part 2: Primary analyses

## 2.1 Load processed data

```{r load-data}
dat <- read.csv("Output/Photo.R_full.csv")
dat$Species <- factor(dat$Species, levels = sp_order)

cat(sprintf("Loaded %d rows, %d individuals, %d species\n",
            nrow(dat), length(unique(dat$individual.ID)),
            length(unique(dat$Species))))
cat(sprintf("Rate column range: %.2f to %.2f μmol O₂ g⁻¹ h⁻¹\n",
            min(dat$umol.cm2.hr), max(dat$umol.cm2.hr)))
cat(sprintf("log.rate = ln(umol.cm2.hr)? %s\n",
            all.equal(log(dat$umol.cm2.hr), dat$log.rate, tolerance = 1e-6)))
```


## 2.2 Q₁₀ analysis (20–32°C)

Q₁₀ calculated from species mean untransformed rates at 20°C and 32°C.

```{r q10-species}
q10_species <- dat %>%
  filter(Temp.Cat %in% c(20, 32)) %>%
  group_by(Species, Temp.Cat) %>%
  summarise(mean_rate = mean(umol.cm2.hr), .groups = "drop") %>%
  pivot_wider(names_from = Temp.Cat, values_from = mean_rate,
              names_prefix = "T") %>%
  mutate(Q10 = (T32 / T20)^(10 / 12))

cat("Species-mean Q10 (20–32°C):\n")
print(as.data.frame(q10_species))
```

```{r q10-individual}
q10_ind <- dat %>%
  filter(Temp.Cat %in% c(20, 32)) %>%
  select(individual.ID, Species, Temp.Cat, umol.cm2.hr) %>%
  pivot_wider(names_from = Temp.Cat, values_from = umol.cm2.hr,
              names_prefix = "T") %>%
  filter(!is.na(T20) & !is.na(T32)) %>%
  mutate(Q10 = (T32 / T20)^(10 / 12))

q10_ind_summary <- q10_ind %>%
  group_by(Species) %>%
  summarise(Q10_mean = mean(Q10), Q10_sd = sd(Q10),
            Q10_se = sd(Q10) / sqrt(n()), n = n(), .groups = "drop")

cat("Individual-level Q10 (mean ± SD):\n")
print(as.data.frame(q10_ind_summary))
```


## 2.3 Arrhenius activation energy (5-point, 20–32°C)

E estimated from linear regressions of ln(untransformed rate) vs. 1/kT
for each replicate, using 5 temperatures (20, 24, 28, 30, 32°C).

```{r arrhenius-regressions}
rising <- dat %>%
  filter(Temp.Cat %in% c(20, 24, 28, 30, 32)) %>%
  mutate(inv_kT = 1 / (k_B * (Temp.C + 273.15)),
         ln_rate = log(umol.cm2.hr))

arr_results <- rising %>%
  group_by(individual.ID, Species) %>%
  summarise(
    n_pts = n(),
    model = list(lm(ln_rate ~ inv_kT)),
    .groups = "drop"
  ) %>%
  mutate(
    E = -map_dbl(model, ~ coef(.x)[2]),
    r2 = map_dbl(model, ~ summary(.x)$r.squared),
    p = map_dbl(model, ~ {
      s <- summary(.x)
      pf(s$fstatistic[1], s$fstatistic[2], s$fstatistic[3], lower.tail = FALSE)
    }),
    slope = map_dbl(model, ~ coef(.x)[2]),
    intercept = map_dbl(model, ~ coef(.x)[1])
  ) %>%
  select(-model)

cat("Individual E values:\n")
print(as.data.frame(arr_results %>% select(individual.ID, Species, E, r2, p)))
```

```{r arrhenius-summary}
arr_summary <- arr_results %>%
  group_by(Species) %>%
  summarise(E_mean = mean(E), E_sd = sd(E), E_se = sd(E) / sqrt(n()),
            r2_mean = mean(r2), n = n(), n_sig = sum(p < 0.05),
            .groups = "drop")

cat("Species E summaries:\n")
print(as.data.frame(arr_summary))
```

```{r arrhenius-anova}
cat("4-species ANOVA on Arrhenius E:\n")
e_aov <- aov(E ~ Species, data = arr_results)
print(summary(e_aov))

# Adjusted R²
s <- summary(e_aov)[[1]]
r2 <- s$`Sum Sq`[1] / sum(s$`Sum Sq`)
n <- sum(s$Df) + 1
adj_r2 <- 1 - (1 - r2) * (n - 1) / (n - s$Df[1] - 1)
cat(sprintf("R² = %.4f, Adj R² = %.4f\n", r2, adj_r2))
```

```{r arrhenius-pairwise}
cat("Pairwise t-tests (Bonferroni):\n")
pw_e <- pairwise.t.test(arr_results$E, arr_results$Species,
                         p.adjust.method = "bonferroni")
print(pw_e)
```

```{r e-q10-correlation}
cat("E–Q10 cross-validation (4 species):\n")
print(cor.test(arr_summary$E_mean, q10_species$Q10))
```


## 2.4 Rate-by-temperature ANOVAs (log-transformed)

```{r rate-anovas}
temps <- c(20, 24, 28, 30, 32, 34, 36)
rate_anova_results <- data.frame()

cat("One-way ANOVA (log-transformed) at each temperature:\n\n")
for (temp in temps) {
  d <- dat %>% filter(Temp.Cat == temp) %>%
    mutate(log_rate = log(umol.cm2.hr))
  aov_res <- aov(log_rate ~ Species, data = d)
  s <- summary(aov_res)
  f_val <- s[[1]]$`F value`[1]
  p_val <- s[[1]]$`Pr(>F)`[1]
  n <- nrow(d)
  ss_mod <- s[[1]]$`Sum Sq`[1]
  ss_res <- s[[1]]$`Sum Sq`[2]
  r2 <- ss_mod / (ss_mod + ss_res)
  adj_r2 <- 1 - (1 - r2) * (n - 1) / (n - s[[1]]$Df[1] - 1)
  sig <- ifelse(p_val < 0.001, "***", ifelse(p_val < 0.01, "**",
         ifelse(p_val < 0.05, "*", "ns")))

  cat(sprintf("  %d°C: F(%d,%d) = %.2f, p = %.6f, Adj R² = %.3f %s\n",
              temp, s[[1]]$Df[1], s[[1]]$Df[2], f_val, p_val, adj_r2, sig))

  rate_anova_results <- rbind(rate_anova_results,
    data.frame(temp = temp, F = f_val, p = p_val,
               adj_r2 = adj_r2, n = n, df_resid = s[[1]]$Df[2]))
}
```

### ANOVA assumption diagnostics

Shapiro-Wilk tests for normality and Levene's tests for homoscedasticity
on log-transformed rates at each temperature, as described in Methods.

```{r assumption-tests}
# ANOVA assumption diagnostics
cat("ANOVA assumption diagnostics (log-transformed rates):\n\n")
cat(sprintf("%-6s | %-24s | %-20s\n",
            "Temp", "Shapiro-Wilk residuals", "Levene's"))
cat(paste(rep("-", 60), collapse = ""), "\n")

for (temp in temps) {

  d <- dat %>%
    filter(Temp.Cat == temp) %>%
    mutate(log_rate = log(umol.cm2.hr))

  # Fit one-way ANOVA
  aov_res <- aov(log_rate ~ Species, data = d)

  # Normality of ANOVA residuals
  sw <- shapiro.test(residuals(aov_res))

  # Homogeneity of variance
  lev <- leveneTest(log_rate ~ Species, data = d)
  lev_p <- lev$`Pr(>F)`[1]

  cat(sprintf("%-6s | p = %-18.3f | p = %.3f\n",
              paste0(temp, "°C"),
              sw$p.value,
              lev_p))
}

# Arrhenius E assumption diagnostics
cat("\nArrhenius E assumption diagnostics:\n")

for (sp in sp_order) {
  sp_e <- arr_results %>%
    filter(Species == sp) %>%
    pull(E)

  sw <- shapiro.test(sp_e)

  cat(sprintf("  %s: Shapiro-Wilk p = %.3f\n",
              sp, sw$p.value))
}

lev_e <- leveneTest(E ~ Species, data = arr_results)

cat(sprintf("  Levene's test: p = %.3f\n",
            lev_e$`Pr(>F)`[1]))
```

```{r habitat-comparisons}
cat("\nHabitat group comparisons (octocorals vs shallow):\n\n")
for (temp in temps) {
  d <- dat %>% filter(Temp.Cat == temp) %>%
    mutate(log_rate = log(umol.cm2.hr))
  octo <- d %>% filter(Species %in% c("Pdarw", "Mplan")) %>% pull(log_rate)
  shallow <- d %>% filter(Species %in% c("Bgran", "Tcocc")) %>% pull(log_rate)
  tt <- t.test(octo, shallow)
  octo_raw <- d %>% filter(Species %in% c("Pdarw", "Mplan")) %>%
    pull(umol.cm2.hr) %>% mean()
  shallow_raw <- d %>% filter(Species %in% c("Bgran", "Tcocc")) %>%
    pull(umol.cm2.hr) %>% mean()
  pct <- ((octo_raw - shallow_raw) / shallow_raw) * 100
  
  sig <- ifelse(tt$p.value < 0.001, "***", ifelse(tt$p.value < 0.01, "**",
       ifelse(tt$p.value < 0.05, "*", "ns")))

p_text <- ifelse(
  tt$p.value < 0.0001,
  "<0.0001",
  sprintf("%.4f", tt$p.value)
)

cat(sprintf(
  "  %d°C: Octo=%.1f, Shallow=%.1f (%+.0f%%), p=%s %s\n",
  temp, octo_raw, shallow_raw, pct, p_text, sig
))

}
```


## 2.5 Site temperature comparison (2023 logger data)

Reproduces the 25.4 vs. 24.7°C comparison reported in Discussion Section 4.1.

```{r temperature-comparison}
temp_logger <- read.csv("Data/temperature_logger_2023.csv")

cat("Site temperature comparison (Feb-Mar 2023):\n\n")
cat(sprintf("  Cerro Tijeretas (shallow): mean = %.1f°C\n",
            mean(temp_logger$tijeretas, na.rm = TRUE)))
cat(sprintf("  Cerro Mundo (deeper):      mean = %.1f°C\n",
            mean(temp_logger$`cerro mundo`, na.rm = TRUE)))

# Date range
date_range_ct <- range(temp_logger$`date tijeretas`, na.rm = TRUE)
date_range_cm <- range(temp_logger$`date cerro mundo`, na.rm = TRUE)
n_days <- as.numeric(difftime(date_range_ct[2], date_range_ct[1], units = "days"))
cat(sprintf("  Recording period: %s to %s (%.0f days)\n",
            format(date_range_ct[1], "%Y-%m-%d"),
            format(date_range_ct[2], "%Y-%m-%d"),
            n_days))
```


---

# Part 3: TPC model selection (untransformed rates, 5 models)

Assesses whether each replicate's thermal response is better described
by a peaked or monotonically increasing model. Five candidates:

- **Peaked:** Gaussian (3 params), Pawar 2018 (4 params), Sharpe-Schoolfield high (4 params)
- **Monotonic:** Simple Arrhenius exponential (2 params), Boltzmann sigmoid (4 params)

The simple exponential has fewer parameters than any peaked model,
ensuring AIC cannot favor peaked models through parameter penalty alone.

```{r model-definitions}
# Peaked models
gaussian_1987 <- function(temp, rmax, topt, a) {
  rmax * exp(-0.5 * ((temp - topt) / a)^2)
}

pawar_2018 <- function(temp, r_tref, e, eh, topt) {
  tref <- 20
  T <- temp + 273.15; Tref <- tref + 273.15; Topt_K <- topt + 273.15
  boltz <- exp(e * (1 / (k_B * Tref) - 1 / (k_B * T)))
  inact_num <- 1 + (e / (eh - e)) * exp((eh / k_B) * (1 / Topt_K - 1 / T))
  inact_den <- 1 + (e / (eh - e)) * exp((eh / k_B) * (1 / Topt_K - 1 / Tref))
  result <- r_tref * boltz * (inact_den / inact_num)
  ifelse(is.finite(result), result, 0)
}

sharpeschoolhigh_1981 <- function(temp, r_tref, e, eh, th) {
  tref <- 20
  T <- temp + 273.15; Tref <- tref + 273.15; Th <- th + 273.15
  boltz <- exp((e / k_B) * (1 / Tref - 1 / T))
  inact <- 1 + exp((eh / k_B) * (1 / Th - 1 / T))
  inact_ref <- 1 + exp((eh / k_B) * (1 / Th - 1 / Tref))
  result <- r_tref * boltz * (inact_ref / inact)
  ifelse(is.finite(result), result, 0)
}

# Monotonic models
simple_exponential <- function(temp, r0, e) {
  tref <- 20
  T <- temp + 273.15; Tref <- tref + 273.15
  r0 * exp((e / k_B) * (1 / Tref - 1 / T))
}

boltzmann_sigmoid <- function(temp, a, b, tm, s) {
  result <- a + (b - a) / (1 + exp(-(temp - tm) / s))
  ifelse(is.finite(result), result, b)
}

# AICc
calc_aicc <- function(mod, n) {
  k <- length(coef(mod))
  AIC(mod) + (2 * k * (k + 1)) / (n - k - 1)
}
```

```{r fit-all-models}
# Model specifications
model_specs <- list(
  Exponential = list(
    formula = umol.cm2.hr ~ simple_exponential(Temp.C, r0, e),
    start = list(r0 = 15, e = 0.5),
    lower = c(1, 0.01), upper = c(80, 3),
    n_params = 2, type = "Monotonic"
  ),
  Gaussian = list(
    formula = umol.cm2.hr ~ gaussian_1987(Temp.C, rmax, topt, a),
    start = list(rmax = 40, topt = 33, a = 5),
    lower = c(0, 20, 0.5), upper = c(100, 45, 30),
    n_params = 3, type = "Peaked"
  ),
  Boltzmann = list(
    formula = umol.cm2.hr ~ boltzmann_sigmoid(Temp.C, a, b, tm, s),
    start = list(a = 10, b = 50, tm = 28, s = 3),
    lower = c(0, 10, 20, 0.1), upper = c(30, 100, 40, 10),
    n_params = 4, type = "Monotonic"
  ),
  Pawar = list(
    formula = umol.cm2.hr ~ pawar_2018(Temp.C, r_tref, e, eh, topt),
    start = list(r_tref = 15, e = 0.5, eh = 2.0, topt = 34),
    lower = c(1, 0.1, 0.5, 25), upper = c(80, 2.0, 10, 42),
    n_params = 4, type = "Peaked"
  ),
  SS_high = list(
    formula = umol.cm2.hr ~ sharpeschoolhigh_1981(Temp.C, r_tref, e, eh, th),
    start = list(r_tref = 15, e = 0.5, eh = 2.0, th = 36),
    lower = c(1, 0.1, 0.5, 30), upper = c(80, 2.0, 10, 45),
    n_params = 4, type = "Peaked"
  )
)

# Fit all models to each replicate
model_results <- data.frame()

for (ind in unique(dat$individual.ID)) {
  d <- dat %>% filter(individual.ID == ind) %>% arrange(Temp.C)
  sp <- as.character(d$Species[1])
  n <- nrow(d)
  
  row <- data.frame(individual_ID = ind, Species = sp, n = n,
                    stringsAsFactors = FALSE)
  aiccs <- c()
  
  for (mname in names(model_specs)) {
    ms <- model_specs[[mname]]
    fit <- tryCatch({
      nlsLM(ms$formula, data = d,
            start = ms$start, lower = ms$lower, upper = ms$upper,
            control = nls.lm.control(maxiter = 500))
    }, error = function(e) NULL)
    
    if (!is.null(fit)) {
      aicc <- calc_aicc(fit, n)
      row[[paste0(mname, "_aicc")]] <- aicc
      aiccs[mname] <- aicc
    } else {
      row[[paste0(mname, "_aicc")]] <- Inf
      aiccs[mname] <- Inf
    }
  }
  
  row$best_model <- names(which.min(aiccs))
  row$best_aicc  <- min(aiccs)
  row$model_type <- model_specs[[row$best_model]]$type
  
  # Delta AICs
  for (mname in names(model_specs)) {
    col <- paste0(mname, "_aicc")
    row[[paste0("delta_", mname)]] <- ifelse(
      is.finite(row[[col]]), row[[col]] - row$best_aicc, NA
    )
  }
  
  model_results <- bind_rows(model_results, row)
}
```

```{r model-selection-results}
cat("Model selection summary:\n\n")
for (sp in sp_order) {
  sp_data <- model_results %>% filter(Species == sp)
  peaked <- sum(sp_data$model_type == "Peaked")
  mono   <- sum(sp_data$model_type == "Monotonic")
  cat(sprintf("  %s: Peaked=%d/%d, Monotonic=%d/%d\n",
              sp, peaked, nrow(sp_data), mono, nrow(sp_data)))
}

total_p <- sum(model_results$model_type == "Peaked")
total_m <- sum(model_results$model_type == "Monotonic")
cat(sprintf("\n  TOTAL: Peaked=%d/%d, Monotonic=%d/%d\n",
            total_p, nrow(model_results), total_m, nrow(model_results)))
cat("\nModel frequency:\n")
print(table(model_results$best_model))
```


---

# Part 4: Figures

## Figure 1: Thermal response profiles

```{r fig1, fig.width=8, fig.height=5.5}
rate_summary <- dat %>%
  group_by(Species, Temp.Cat) %>%
  summarise(mean_rate = mean(umol.cm2.hr),
            se_rate = sd(umol.cm2.hr) / sqrt(n()),
            .groups = "drop")

fig1 <- ggplot() +
  geom_jitter(data = dat,
              aes(x = Temp.Cat, y = umol.cm2.hr, color = Species),
              alpha = 0.25, size = 1.5, width = 0.3) +
  geom_smooth(data = rate_summary,
              aes(x = Temp.Cat, y = mean_rate, color = Species, group = Species),
              method = "loess", span = 1.5, se = FALSE, linewidth = 1.2) +
  geom_errorbar(data = rate_summary,
                aes(x = Temp.Cat, y = mean_rate,
                    ymin = mean_rate - se_rate, ymax = mean_rate + se_rate,
                    color = Species),
                width = 0.5, linewidth = 0.8) +
  geom_point(data = rate_summary,
             aes(x = Temp.Cat, y = mean_rate, color = Species), size = 3) +
  scale_color_manual(values = sp_colors,
                     breaks = sp_order,
                     labels = c(expression(italic("P. darwinii")),
                                expression(italic("M. plantaginea")),
                                expression(italic("B. grande")),
                                expression(italic("T. coccinea")))) +
  labs(x = "Temperature (°C)",
       y = expression(Respiration~rate~(mu*mol~O[2]~g^{-1}~h^{-1}))) +
  scale_x_continuous(breaks = c(20, 24, 28, 30, 32, 34, 36, 38)) +
  scale_y_continuous(limits = c(5, 66.5), breaks = seq(10, 60, by = 10),
                     expand = c(0, 0)) +
  theme_classic(base_size = 14) +
  theme(legend.position = "top", legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.margin = margin(b = -5)) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

fig1
ggsave("Figures/Fig1_thermal_response_profiles.pdf", fig1, width = 8, height = 5.5)
```


## Figure 2: Temperature sensitivity

```{r fig2, fig.width=10, fig.height=5}
species_labels <- c(
  "Pdarw" = expression(italic("P. darwinii")),
  "Mplan" = expression(italic("M. plantaginea")),
  "Bgran" = expression(italic("B. grande")),
  "Tcocc" = expression(italic("T. coccinea"))
)

# Panel (a): Q10
p2a <- ggplot(q10_species %>% mutate(Species = factor(Species, levels = sp_order)),
              aes(x = Species, y = Q10, fill = Species)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.2f", Q10)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = sp_colors) +
  scale_x_discrete(labels = species_labels) +
  scale_y_continuous(limits = c(0, 3), expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = expression(italic(Q)[10]~"(20–32°C)"), title = "(a)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none", plot.title = element_text(face = "bold", size = 14))

# Panel (b): Arrhenius E
sig_letters <- arr_results %>%
  mutate(Species = factor(Species, levels = sp_order)) %>%
  group_by(Species) %>%
  summarise(y = max(E) + 0.04, .groups = "drop") %>%
  mutate(label = c("a", "b", "c", "c"))

p2b <- ggplot(arr_results %>% mutate(Species = factor(Species, levels = sp_order)),
              aes(x = Species, y = E)) +
  geom_jitter(aes(color = Species), width = 0.10, alpha = 0.55, size = 2.5) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.18, linewidth = 0.7, color = "black") +
  stat_summary(fun = mean, geom = "point", size = 3, color = "black") +
  geom_text(data = sig_letters, aes(x = Species, y = y, label = label),
            inherit.aes = FALSE, fontface = "bold", size = 5) +
  scale_color_manual(values = sp_colors) +
  scale_x_discrete(labels = species_labels) +
  labs(x = NULL, y = expression("Activation energy, " * italic(E) * " (eV)"),
       title = "(b)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none", plot.title = element_text(face = "bold", size = 14))

fig2 <- plot_grid(p2a, p2b, ncol = 2, align = "hv", axis = "tb")
fig2
ggsave("Figures/Fig2_temperature_sensitivity.pdf", fig2, width = 10, height = 5)
```


## Figure 3: Arrhenius regressions

```{r fig3, fig.width=7, fig.height=5.5}
arr_lines <- arr_results %>%
  group_by(Species) %>%
  summarise(mean_slope = mean(slope), mean_int = mean(intercept),
            mean_E = mean(E), .groups = "drop")

inv_kT_range <- rising %>%
  group_by(Species) %>%
  summarise(min_ikT = min(inv_kT), max_ikT = max(inv_kT), .groups = "drop")

pred_lines <- arr_lines %>%
  left_join(inv_kT_range, by = "Species") %>%
  rowwise() %>%
  mutate(data = list(tibble(
    inv_kT = seq(min_ikT, max_ikT, length.out = 100),
    ln_rate = mean_slope * seq(min_ikT, max_ikT, length.out = 100) + mean_int
  ))) %>%
  unnest(data)

temp_breaks_C <- c(20, 24, 28, 32)

legend_labels <- arr_lines %>%
  arrange(factor(Species, levels = sp_order)) %>%
  mutate(lab = paste0("<i>", sp_names[Species],
                      "</i> (<i>E</i> = ", round(mean_E, 2), " eV)")) %>%
  pull(lab)

fig3 <- ggplot() +
  geom_point(data = rising,
             aes(x = inv_kT, y = ln_rate, color = Species),
             alpha = 0.30, size = 1.5) +
  geom_line(data = pred_lines,
            aes(x = inv_kT, y = ln_rate, color = Species), linewidth = 1.4) +
  scale_color_manual(values = sp_colors, labels = legend_labels) +
  scale_x_reverse(
    name = expression("1/kT (" * eV^{-1} * ")"),
    sec.axis = sec_axis(~ 1 / (k_B * .) - 273.15,
                        name = "Temperature (°C)", breaks = temp_breaks_C)
  ) +
  labs(y = "ln(Respiration rate)") +
  theme_classic(base_size = 14) +
  theme(legend.position = c(0.75, 0.15), legend.title = element_blank(),
        legend.text = ggtext::element_markdown(size = 10),
        legend.background = element_rect(fill = alpha("white", 0.90)))

fig3
ggsave("Figures/Fig3_arrhenius_regressions.pdf", fig3, width = 7, height = 5.5)
```


## Figure S1: Individual profiles with model fits

Uses the best model determined by the 5-model comparison in Part 3 
(`model_results`). For plotting, we refit only the winning model type 
(Exponential or Gaussian) to generate smooth prediction curves.

```{r figS1-fit, fig.width=10, fig.height=8}

# Generate predictions using the best model from model_results (Part 3)
pred_list <- list()

for (ind in unique(dat$individual.ID)) {

  d  <- dat %>% filter(individual.ID == ind) %>% arrange(Temp.C)
  sp <- as.character(d$Species[1])
  n  <- nrow(d)
  
  # Look up the best model from the 5-model selection
  best_model <- model_results %>% 
    filter(individual_ID == ind) %>% 
    pull(best_model)
  best_type <- model_results %>% 
    filter(individual_ID == ind) %>% 
    pull(model_type)

  pred_temps <- seq(min(d$Temp.C), max(d$Temp.C), by = 0.5)

  # Refit the winning model type to generate prediction curve
  pred_rates <- if (best_type == "Monotonic") {
    fit <- tryCatch(
      nlsLM(umol.cm2.hr ~ simple_exponential(Temp.C, r0, e),
            data = d, start = list(r0 = 15, e = 0.5),
            lower = c(1, 0.01), upper = c(80, 3),
            control = nls.lm.control(maxiter = 200)),
      error = function(e) NULL)
    if (!is.null(fit)) predict(fit, newdata = data.frame(Temp.C = pred_temps))
    else rep(NA_real_, length(pred_temps))
  } else {
    fit <- tryCatch(
      nlsLM(umol.cm2.hr ~ gaussian_1987(Temp.C, rmax, topt, a),
            data = d, start = list(rmax = 40, topt = 33, a = 5),
            lower = c(0, 20, 0.5), upper = c(100, 45, 30),
            control = nls.lm.control(maxiter = 200)),
      error = function(e) NULL)
    if (!is.null(fit)) predict(fit, newdata = data.frame(Temp.C = pred_temps))
    else rep(NA_real_, length(pred_temps))
  }

  pred_list[[ind]] <- data.frame(
    individual.ID = ind,
    Species       = sp,
    model_type    = best_type,
    Temp.C        = pred_temps,
    rate          = pred_rates,
    stringsAsFactors = FALSE
  )
}

pred_df <- bind_rows(pred_list)
pred_df$Species <- factor(pred_df$Species, levels = sp_order)

cat("Fig S1 model types (from Part 3 five-model selection):\n")
print(table(model_results$Species, model_results$model_type))

figS1 <- ggplot() +
  geom_point(data = dat,
             aes(x = Temp.Cat, y = umol.cm2.hr, color = Species),
             alpha = 0.4, size = 1) +
  geom_line(data = pred_df,
            aes(x = Temp.C, y = rate, group = individual.ID, linetype = model_type),
            alpha = 0.6, linewidth = 0.5) +
  facet_wrap(~ Species, scales = "free_y",
             labeller = labeller(Species = sp_names)) +
  scale_color_manual(values = sp_colors, guide = "none") +
  
  scale_linetype_manual(values = c("Monotonic" = "solid", "Peaked" = "dashed"),
                      name = "Best-fit model",
                      labels = c("Monotonic" = "Exponential", "Peaked" = "Gaussian")) +
  labs(x = "Temperature (°C)",
       y = expression(Respiration~rate~(mu*mol~O[2]~g^{-1}~h^{-1}))) +
  theme_classic(base_size = 12) +
  theme(strip.text = element_text(face = "italic"), legend.position = "bottom")

figS1
ggsave("Figures/FigS1_individual_profiles.pdf", figS1, width = 10, height = 8)
```


---

# Part 5: Export tables

```{r export-tables}
# Table S1: Individual E and Q10
table_s1 <- arr_results %>%
  select(individual.ID, Species, E, r2, p, n_pts) %>%
  left_join(q10_ind %>% select(individual.ID, Q10), by = "individual.ID") %>%
  arrange(Species, individual.ID)
write.csv(table_s1, "Output/Table_S1_individual_E_Q10.csv", row.names = FALSE)

# Table S2: Pairwise E comparisons (expanded format)
pw_expanded <- data.frame()
for (i in 1:(length(sp_order)-1)) {
  for (j in (i+1):length(sp_order)) {
    sp1 <- sp_order[i]; sp2 <- sp_order[j]
    g1 <- arr_results %>% filter(Species == sp1) %>% pull(E)
    g2 <- arr_results %>% filter(Species == sp2) %>% pull(E)
    tt <- t.test(g1, g2, var.equal = TRUE)
    pw_expanded <- rbind(pw_expanded, data.frame(
      Species1 = sp1, Species2 = sp2,
      mean_diff = round(mean(g1) - mean(g2), 3),
      t = round(tt$statistic, 3),
      p_bonferroni = round(min(tt$p.value * 6, 1), 6)
    ))
  }
}
write.csv(pw_expanded, "Output/Table_S2_pairwise_E.csv", row.names = FALSE)


# Table S3: Full 5-model AIC selection
table_s3 <- model_results %>%
  select(individual_ID, Species, n,
         delta_Exponential, delta_Gaussian, delta_Boltzmann,
         delta_Pawar, delta_SS_high,
         best_model, model_type) %>%
  mutate(across(starts_with("delta_"), ~ round(., 1)))
write.csv(table_s3, "Output/Table_S3_model_selection.csv", row.names = FALSE)

# ============================================================
# Main manuscript Table 1: Species summary
# ============================================================

# Respiration mean ± SD at representative temperatures
rate_table1 <- dat %>%
  filter(Temp.Cat %in% c(20, 28, 32)) %>%
  group_by(Species, Temp.Cat) %>%
  summarise(
    mean_rate = mean(umol.cm2.hr, na.rm = TRUE),
    sd_rate   = sd(umol.cm2.hr, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    rate_sd = sprintf("%.1f ± %.1f", mean_rate, sd_rate)
  ) %>%
  select(Species, Temp.Cat, rate_sd) %>%
  pivot_wider(
    names_from = Temp.Cat,
    values_from = rate_sd,
    names_prefix = "Rate_"
  )

# Match Q10 and E summaries by Species rather than row position
table_1 <- tibble(Species = sp_order) %>%
  left_join(
    q10_species %>% select(Species, Q10),
    by = "Species"
  ) %>%
  left_join(
    as.data.frame(arr_summary) %>%
      select(Species, E_mean, E_sd),
    by = "Species"
  ) %>%
  left_join(rate_table1, by = "Species") %>%
  mutate(
    Species = sp_names[as.character(Species)],
    n = 9,
    Q10 = round(Q10, 2),
    E_mean_sd = sprintf("%.3f ± %.3f", E_mean, E_sd)
  ) %>%
  select(
    Species, n, Q10, E_mean_sd,
    Rate_20, Rate_28, Rate_32
  )

write.csv(
  table_1,
  "Output/Table_1.csv",
  row.names = FALSE
)

# ============================================================
# Main manuscript Table 2: Model summaries
# ============================================================

# Pull Arrhenius E statistics directly from the fitted ANOVA
e_tab <- summary(e_aov)[[1]]

e_F <- e_tab$`F value`[1]
e_p <- e_tab$`Pr(>F)`[1]
e_df_resid <- e_tab$Df[2]

# Calculate R2 and adjusted R2 using the same approach used above
e_ss_model <- e_tab$`Sum Sq`[1]
e_ss_resid <- e_tab$`Sum Sq`[2]
e_r2 <- e_ss_model / (e_ss_model + e_ss_resid)

e_n <- nrow(arr_results)
e_df_model <- e_tab$Df[1]

e_adj_r2 <- 1 - (1 - e_r2) * (e_n - 1) /
  (e_n - e_df_model - 1)

e_row <- data.frame(
  Metric   = "Arrhenius E",
  df_resid = e_df_resid,
  n        = e_n,
  adj_r2   = round(e_adj_r2, 3),
  F        = round(e_F, 2),
  p        = ifelse(
    e_p < 0.001,
    "< 0.001",
    sprintf("%.3f", e_p)
  )
)

# Temperature-specific rate ANOVAs generated earlier
rate_rows <- rate_anova_results %>%
  transmute(
    Metric   = paste0("Rate at ", temp, "°C"),
    df_resid = df_resid,
    n        = n,
    adj_r2   = round(adj_r2, 3),
    F        = round(F, 2),
    p        = ifelse(
      p < 0.001,
      "< 0.001",
      sprintf("%.3f", p)
    )
  )

table_2 <- bind_rows(e_row, rate_rows)

write.csv(
  table_2,
  "Output/Table_2.csv",
  row.names = FALSE
)

# Rate ANOVA summary
write.csv(rate_anova_results, "Output/Table_rate_ANOVAs.csv", row.names = FALSE)

# E species summary
write.csv(as.data.frame(arr_summary), "Output/Table_E_summary.csv", row.names = FALSE)

cat("All tables exported to Output/\n")
```


---

# Session info

```{r session-info}
sessionInfo()
```
